Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden

### 2023-09-13 analysis/paper/paper.qmd
Updated manuscript

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Environmental Archaeology Lab, Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  pdf:
    documentclass: scrartcl
    fig-pos: "H"
    geometry: margin = 2.54cm 
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials. 125 points and preforms made from quartz/quartzite were sampled from 46 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Hunter-Gatherer; Lithics; Provenance; Quartzite; Vibrational spectroscopy; Scandinavia
---

@@

Petrographic characterization and provenance studies are instrumental for inferring trade and exchange among prehistoric societies [@tykot03]. The presence of lithic material foreign to the local region of an archaeological site, raises important questions on the processes that resulted in its eventual deposition and discovery. The characterization of geological material and identification of geological sources are the first steps in reconstructing a prehistoric community’s strategies for raw material selection and management. Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17]. This focus has led to a bias, however, which has been highlighted by Prieto et al. [-@prieto_etal19: 15]. Recent decades, however, have seen an increase in interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].

@@

In this study, three different spectroscopic techniques have been applied to material stored in collections at Västerbottens museum in Umeå, Sweden. X-Ray Fluorescence (XRF) is a well-established spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data. The use of near-infrared spectroscopy (NIRS) is similarly well-known, as described by Norris [-@norris96], and there is a large body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].

# Aims {-}

This paper presents an assessment of the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources. The geological complexity of both the region and the material necessitates alternative approaches to studies of raw material management. Without relying on well-delimited source areas for geological material, a predictive model capable of separating different “classes” of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models. The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden between 4 000 and 2 000 BP.

@@

This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of quartz and quartzite. The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map & @tbl-point-data). The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and XRF.

The geographically closest chronology for bifacial points is centred around southeast Norway [@mjaerum12], but a few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89]. The latter suggest a period of use between ca. 3 800 - 2 000 BP [@kaeck09: 60]. Studies of raw material use at sites dating to this period seem to indicate an increase in the use of quartzite when compared to earlier periods [@baudou78].

\linebreak
<!--
```{r}
#| label: fig-distribution-map
#| cache: false
#| cache-lazy: false
#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden. The glacial disturbance is visible in the widely distributed moraine. Tectonic symbology after [@gee_stephens20]. Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet"

knitr::include_graphics(here::here("analysis", "figures", "site_distribution_v1.jpg"), error = FALSE)
```
-->

![Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden. The glacial disturbance is visible in the widely distributed moraine. Tectonic symbology after Gee and Stephens (2020). Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet](../figures/site_distribution_v1.jpg){#fig-distribution-map width=100%, height=100%}

@@

The dataset includes three main groups of petrographic material, as classified by archaeologists: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Artefacts that had no prior classification were sorted into either the quartz or quartzite category based on visual inspection. This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated.

In his analysis of the petrographic material in the Early Norrland project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91: 24]. As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.

\linebreak
<!--
```{r}
#| label: fig-artefacts
#| cache: false
#| cache-lazy: false
#| fig-cap: "Examples of the bifacial points studied. The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm"

knitr::include_graphics(here::here("analysis", "figures", "figure_2.jpg"), error = FALSE)
```
-->

![Examples of the bifacial points studied. The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm](../figures/figure_2.jpg){#fig-artefacts width=100%, height=100%}

@@

```{r}
#| label: tbl-point-data
#| cache: false
#| cache-lazy: false
#| tbl-cap: "Table of artefact location, material and type"

@@

All measurements were collected at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/). The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being 2 computer screens. The sampled material is heterogeneous in colour and shape (@fig-artefacts - @fig-artefact-size) (see supplementary data). The artefacts range from small and flat point fragments, to large and rough preforms. Therefore, the signal integration could at times be difficult, leading to scattering effects and saturation. For NIR and Raman analyses, each artefact was sampled twice, on different sides, with additional measurements where sampling was difficult. When sampling the probe was in contact with the surface of the artefact at as flat a surface as possible. If the material featured high heterogeneity, the sample point was focused on the surrounding matrix, avoiding deviating inclusions (e.g. the dark mineral veins in @fig-artefact-size).

For XRF, the more time-consuming nature of the measurement process necessitated only one measurement per object. The representativeness of the XRF analysis is thus lower, and inferences derived from the results in this study should be used with caution. The risk of misinterpretation is, however, somewhat compensated by the large size of the dataset, with larger sites including multiple objects that may come from the same geological source. The results from sites with smaller assemblages are, however, more liable to be affected by the limited range of material analysed.

@@

@@

Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near-infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify it. Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77: 508]. Overtones occur at frequencies that are a multiple of the infrared vibration, and most of these end up in the near-infrared region [@osborne_etal93].

NIR analysis was conducted with an Analytical Spectral Device (ASD) LabSpec 4, using a contact probe (spot size: 10 mm). The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before Principal Component Analysis (PCA) modelling.

@@

### Raman {-}
Raman spectroscopy shares similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared region [@nafie01]. Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^, with a resolution of 9.5 cm^-1^. Integration time was set to 20 seconds with 30% power. After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data. This was achieved using Lieber and Mahadevan-Jansens' [-@lieber_mahadevan-jansen03] algorithm for polynomial fitting, with the ChemoSpec package [@hanson16]. The spectra were then mean-centred before PCA.

@@

### X-Ray Fluorescence (XRF) {-}
XRF spectroscopy is a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Spectral features in XRF spectrometry can be related to certain elements according to Moseley’s law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24]. Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12: 143-159]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.
XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand. The reference calibration used for element quantification was mining mode [@knight_etal21: 2]. The integration time was set to 130 seconds.

As the variables in XRF analysis are given as proportions of a constant sum (e.g. percentage, parts per million (ppm)), this creates something referred to as the closed sum effect [@pawlowsky-glahn_egozcue06]. This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20]. This has implications for classical statistical methods and is managed by transforming the data using log-ratio transformation [@aitchison99]. Currently there are three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15]. In the PCA, the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g. @reimann_etal12]. With this transformation, PCA is thus suitable for exploring geochemical trends related to the rock-forming process without being impacted by the closed sum effect.

As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or “missing values”) during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Missing values are not permitted in the PCA and thus had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15]. *lrEM* stands for “Log-ratio Expectation-Maximisation”, which is an algorithm used to impute “left-censored data” (e.g. values below LOD, rounded zeros). Following log-transformation the data was mean-centred and scaled to unit variance before PCA. Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9; @lundblad_etal08].

@@

The data were evaluated using PCA, an unsupervised exploratory technique for reducing the dimensionality of a dataset  [@geladi_linderholm20]. Multivariate approaches, such as PCA, have been proven useful for exploring complex spectroscopic datasets [@sciuto18]. Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material. These groupings can then be followed up with supervised methods, and aid in the construction of predictive models for classifying unknown materials according to previously defined groups.

PCA analysis was performed using the base R *prcomp()* function [@Rteam21a; @Rteam21b].

@@

Significant peaks were identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the result of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (@fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occurring in the 2 000 nm - 2 500 nm region. While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH [@hunt77: 509; @clark_etal90: 12664]. In the current dataset the 2 200 nm band is typically the more intense band.

@@

The loading lines for the PCA can be seen in the supplementary data. The first principal component (PC 1) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (@fig-nir-pca). This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative. A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %). Objects with a white or light grey hue gravitate towards the bottom left of the score plot, while material that displays some translucence cluster towards the top left. There are objects that break this trend, however, such as sample 255 in the bottom right of @fig-nir-pca :A.

@@

## Raman {-}
There are two distinguishing features present in the Raman data (@fig-raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures. The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, , are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].

The second feature shows as two peaks at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominates the spectra in terms of intensity for the darker samples. These positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures, whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21]. The G band peak falls within the expected position, whereas the position of the D band is more commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behaviour of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07: 1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

>>>>

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Environmental Archaeology Lab, Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  pdf:
    documentclass: scrartcl
    fig-pos: "H"
    geometry: margin = 2.54cm 
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the period 4 000 – 2 000 BP, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials, and provenance of locally available geological material is complex and fraught with uncertainty. 126 points and preforms made from quartz/quartzite were sampled from 47 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Hunter-Gatherer; Lithics; Provenance; Quartzite; Vibrational spectroscopy; Scandinavia
---

@@

Petrographic characterization and provenance studies are instrumental for inferring trade and exchange among prehistoric societies [@tykot03]. The presence of lithic material foreign to the local region of an archaeological site, raises important questions on the processes that resulted in its eventual deposition and discovery. The characterization of geological material and identification of geological sources are the first steps in reconstructing a prehistoric community’s strategies for raw material selection and management. Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17]. This focus has led to a bias which has been highlighted by Prieto et al. [-@prieto_etal19: 15]. Recent decades, however, have seen an increase in interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].

@@

In this study, three different spectroscopic techniques have been applied to material stored in collections at Västerbottens museum in Umeå, Sweden. X-Ray Fluorescence (XRF) is a well-established spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data. The use of near-infrared spectroscopy (NIRS) is similarly well-known, as described by Norris [-@norris96], and there is a large body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, and the Earth sciences [@dubessy_etal12].

The geological complexity of both the region and the material necessitates new approaches to studies of both raw materials and artefact distributions. This paper presents an assessment of the potential for non-destructive spectroscopy to advance the characterization of quartz and quartzite material without the use of references from known geological sources. The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden are limited, however, and thus this aims to:

* Assess the potential of multispectral non-destructive analysis of quartzite using NIRS, Raman and ED-XRF 
*	Assess the variability, and identify features, in quartzite that may be of use in its characterization
*	Discuss the implications of these methods for developing future strategies for studying raw material management of quartzite between 4 000 and 2 000 BP in Northern Sweden

@@

This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of, what has been classified by archaeologists as, brecciated quartz and quartzite. The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map & @tbl-point-data). The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and ED-XRF (@fig-artefacts; @fig-artefact-size).

The geographically closest chronology for bifacial points is centred around southeast Norway [@mjaerum12], but a few attempts have been made for Northern Sweden [@baudou92: 99; @forsberg85: 5; @forsberg89]. The latter suggest a period of use between ca. 3 800 – 2 000 BP [@kaeck09: 60]. Studies of raw material use at sites dating to this period seem to indicate an increase in the use of quartzite when compared to earlier periods [@baudou78].

@@

![Geographical distribution of the sampled sites, with find context, along the river of Ångermanälven, Västerbotten, Sweden. The Late Quaternary reincorporation of geological material from the Scandian mountains into the sediments in the forest inlands adds to the complexity of questions of provenance. Tectonic symbology after (Gee and Stephens 2020). Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.](../figures/Fig1.jpg){#fig-distribution-map width=100%, height=100%}

@@

```{r}
#| label: tbl-point-data
#| cache: false
#| cache-lazy: false
#| tbl-cap: "List of artefact location, material and type."

@@

The dataset includes two main groups of petrographic material, as classified by archaeologists: quartzite and brecciated quartz (Swe. "*brecciekvarts*") (@fig-artefacts; @tbl-point-data). Geological definitions of quartzite include characteristics such as the material being exceptionally hard, predominantly consisting of quartz, and with conchoidal fracturing [@skolnick65; @howard05; @prieto_etal19]. A reliable classification would thus be based on a combination of chemical and physical (macroscopic and microscopic) properties. Archaeologists, however, tend to classify the material in the field on the basis of visual characteristics and perceived hardness. This can lead to considerable ambiguity, but in general, a material will be archaeologically classified as quartzite in Northern Sweden if it is an exceptionally hard, fine-grained, quartz-rich rock.

![Examples of the bifacial points studied. The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm](../figures/Fig2.jpg){#fig-artefacts width=100%, height=100%}

This study includes `r length(quartzite)` and `r length(bquartz)` brecciated quartz samples. The inclusion of brecciated quartz in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated. The term  *brecciated quartz* (also referred to as *mylonite quartz*) was introduced to archaeologists in Northern Sweden by the geologist Åhman [-@ahman67: 8] in his analysis of the petrographic material in the *Early Norrland* project [@biornstad68; @ahman67]. Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis. No detailed information on these microscopic features were mentioned in his original publication. Brecciated quartz in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material that functionally behaves similarly to quartzite [e.g. @holm91: 24] (@fig-artefacts). As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.

@@

All measurements were collected at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/). The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being 2 computer screens. The sampled material is heterogeneous in colour and shape (@fig-artefacts - @fig-artefact-size) (see supplementary data). The artefacts range from small and flat point fragments, to large and rough preforms. This variation led to scattering effects and saturation which at times made signal integration difficult.

For NIR and Raman analyses, each artefact was sampled twice, on different sides, with additional measurements where sampling was difficult. When sampling, the probe was positioned in contact with as flat a surface as possible of the artefact. If the material featured high heterogeneity, the sample point was focused on the surrounding matrix, avoiding deviating inclusions (e.g. the dark mineral veins in @fig-artefact-size).

For XRF, the more time-consuming nature of the measurement process necessitated only one measurement per object. The representativeness of the XRF analysis is thus lower, and inferences derived from the results in this study should be used with caution. The risk of misinterpretation is, however, somewhat compensated for by the large size of the dataset, with larger sites including multiple objects that may come from the same geological source. The results from sites with smaller assemblages are, however, more liable to be affected by the limited range of material analysed.

@@

On the basis of the theory of molecular vibration and excitation [@hunt77], it is possible to generate overtone vibrations when exposing a geological material to electromagnetic radiation. These overtones can be related to certain molecular groups, the hydroxyl (OH) group being the most common one capable of generating them. In quartz-based materials this typically relates to molecular water and metal-OH combinations [@hunt77: 508]. Overtones occur at frequencies that are multiples of the infrared vibration, and most of these end up in the near-infrared region [@osborne_etal93].

NIR analysis was conducted with an Analytical Spectral Device (ASD) LabSpec 4, using a contact probe (spot size: 10 mm). The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling of 1.4 nm in the visible and near-infrared range and 1.1 nm in the short‐wave infrared range. Respectively, spectral resolution is 3 nm in the visible and near-infrared range, and 10 nm in the short-wave infrared range.  Each object was measured against a white reference background and the collected measurements were averaged for each artefact. The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before Principal Component Analysis (PCA) modelling.

@@

### Raman {-}
Raman spectroscopy shares similarities with NIRS in that they both can be used to infer molecular structures based on their vibrational properties within the near-infrared region [@nafie01]. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both *Rayleigh* and *Raman scatter*, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material[@smith_carabatos-nedelec01].

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser (spot size: 85 $\mu$m). The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^, with a resolution of 9.5 cm^-1^. Integration time was set to 20 seconds with 30% power. After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data. This was achieved using Lieber and Mahadevan-Jansens' [-@lieber_mahadevan-jansen03] algorithm for polynomial fitting, with the ChemoSpec package [@hanson16]. The spectra were then mean-centred before PCA.

@@

### X-Ray Fluorescence (XRF) {-}
XRF spectroscopy is a well-established method for petrographic characterization in archaeology [@hall60; @shackley11a]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample [@sitko_zawisa12: 143-159]. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated, as was done in this project.

Measurements were collected using the Energy Dispersive (ED) Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand. The reference calibration used for element quantification was mining mode [@knight_etal21: 2]. The integration time was set to 130 seconds. Higher spectral resolution can be achieved if used with a Wavelength Dispersive (WD) XRF system [@menne_etal20]. However, these are limited to lab-based environments and not applicable in the field, which is central to the current study.

As the variables in XRF analysis are given as proportions of a constant sum (e.g. percentage, parts per million (ppm)), this creates something referred to as the *closed sum* effect [@pawlowsky-glahn_egozcue06]. This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20]. This has implications for classical statistical methods and is managed by transforming the data using log-ratio transformation [@aitchison99].Currently there are three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15]. In the PCA, the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g. @reimann_etal12]. With this transformation, PCA is thus suitable for exploring geochemical trends related to the rock-forming process without being impacted by the closed sum effect.

As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or “missing values”) during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Missing values are not permitted in the PCA and thus had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15]. *lrEM* stands for “Log-ratio Expectation-Maximisation”, which is an algorithm used to impute “left-censored data” (e.g. values below LOD, rounded zeros). Following log-transformation the data was mean-centred and scaled to unit variance before PCA. To prevent quantification errors related to samples size artefacts that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9; @lundblad_etal08].

@@

The data were evaluated using PCA, an unsupervised exploratory technique for reducing the dimensionality of a dataset [@geladi_linderholm20]. Analysis was performed using the base R *prcomp()* function [@Rteam21a; @Rteam21b].

@@

Significant peaks were identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the result of overtones and vibration combinations of OH groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water  (@fig-nir-peak-summary) [@hunt77: 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occurring in the 2 000 nm - 2 500 nm region. While there are exceptions, a more intense 2 200 nm band has typically been related to aluminium hydroxide (AlOH), and a more intense 2 350 nm band to magnesium hydroxide (MgOH) [@hunt77: 509; @clark_etal90: 12664]. In the current dataset the 2 200 nm band is typically the more intense band.

@@

The loading lines for the PCA can be seen in the supplementary data. The first principal component (PC 1) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (@fig-nir-pca). This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative. A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %).Objects with a white or light grey hue gravitate towards the bottom left of the score plot, while material that displays some translucence cluster towards the top left. There are objects that break this trend, however, such as sample 255 in the bottom right of @fig-nir-pca :A.

@@

## Raman {-}
There are two distinguishing features present in the Raman data (@fig-raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures. The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].

second feature shows as two peaks at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominates the spectra in terms of intensity for the darker samples. These positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99]. The D band occurs as a result of disordered C structures, whilst the G band is indicative of sp^2^ C networks and caused by stretching in the C bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21]. The G band peak falls within the expected position, whereas the position of the D band is more commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behaviour of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07: 1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.
